Safe Transferable Regions

### 2017-01-13 paper/type-system.tex
Sec.3 made more peppy.

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$.  We now formally develop this
intuition via \fbname (\FB), our explicitly typed core language (with
region types) that incorporates the features introduced in the
previous section. \fbname builds on the Featherweight Generic Java
(FGJ)~\cite{fgj} formalism, and reuses notations and various
definitions from~\cite{fgj}, such as the definition of type
well-formedness for the core (region-free) language.  (The language
used in Section~\ref{sec:overview} is essentially a version of FGJ
without region types and with some syntactic sugar.)

% In this section, we formally develop \fbname, a language that
% incorporates the features introduced in the previous section, as well
% as its type system, building on the Featherweight Generic Java
% (FGJ)~\cite{fgj} formalism.  Our development reuses notations and
% various definitions from~\cite{fgj}, such as the definition of type
% well-formedness for the core (region-free) language, which are
% elaborated further in the appendix~\cite{techrep}.

@@

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$. Class fields are not allowed to
be type polymorphic, in keeping with C\#, which constitutes the core
of \name. Separate parameterization over types and regions is needed
in \FB to support region-polymorphic higher-order functions as class
fields. For such higher-order class fields, regions are parameterized
at the (class) field level whereas types are parameterized at the
class level. This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
parameterization separate from region parameterization allows \FB to
type such examples. This choice also simplifies type inference in the
presence of higher-order functions.

@@

Like classes, methods can also exhibit constrained region
polymorphism.  A method definition in \FB is necessarily polymorphic
over its allocation context (\S~\ref{sec:alloc-ctxt}), and
optionally polymorphic with respect to the regions containing its
arguments.  Region parameters, like those on classes, are qualified
with constraints ($\phi$).
% Region parameters on the methods, like those on classes, are
% accompanied by constraints ($\phi$) capturing the conditions that the
% parameters need to satisfy for the method to be considered well-formed
% (Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
% or $\rho^a_m$) is the first and inevitable region parameter of every
% method in \FB. 
If a method is not intended to be polymorphic with respect to its
allocation context (for example, if its allocation context needs to be
same as the allocation region of its \emph{this} argument), then the
required monomorphism can be captured as an equality constraint in
$\phi$.

% Like methods, functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
\FB extends FGJ's expression language with a lambda expression and an
application expression ($e\inang{\rbar}(\bar{e})$) to define
and apply functions (lambdas). Functions, like methods, exhibit
constrained region polymorphism, as evident in their arrow region type
($\inang{\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
\tau$).
% at the prenex position, similar to ML type schemes. However,
% unlike ML, we don't distinguish between types and type schemes; any of
% the $\tau$'s in the arrow type can themselves be region-parametric
% arrow types. Like with methods, region parameters on functions can be
% instantiated when they are applied.
% In this respect, our region type system is more like
% System F's type system, which admits higher-rank parametric
% polymorphism. Like System F, \FB provides a region instantiation
% expression ($e\inang{\ralloc\rbar}$), and a region generalization
% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
% instantiate and generalize region variables. 
Since a function closure can escape the context in which it is
created, it is important to keep track of the region in which it is
created in order to avoid unsafe dereferences. The $\rgn$ annotation
above the arrow in the arrow type denotes the allocation region of the
corresponding closure. 
% Note that it is important to distinguish
% between the allocation context argument ($\rhoalloc$) of a function
% and the allocation region ($\rgn$) of its closure. In \name, the later
% corresponds to the region where a \C{Func} object is allocated, while
% the former corresponds to the region where it is applied
% For instance, in the following example:
% \begin{codejava}
% letregion $\rgn$ {
%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
%   in f
% }
% \end{codejava}
% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
% in $\rgn$, and (b). when executed under an allocation context
% $\rho^a$, the closure returns an object allocated in $\rho^a$.

@@

@@

The type rules distinguish between the \C{new} expressions that create
objects of  the \RgnZ class, and \C{new} expressions that create
objects of other classes. The rule for the latter relies on an
auxiliary definition called $\fields$ (undefined for $\RgnZ$ class)
that returns the sequence of type bindings for fields (instance
variables) of a given class type.  Like in FGJ, the names and types of
a constructor's arguments in \FB are same as the names and types of
its class's fields, and the type rule relies on this fact to typecheck
constructor applications.

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\pi \notin \Delta$). This condition is needed in order
to prevent the new region from incorrectly assuming existing outlives
relationships on an eponymous region. Provided this is satisfied, the
expression ($e$) under \C{letregion} is then typechecked assuming that
the new region is live ($\Delta \cup \{\pi\}$) and that it is outlived
by all existing live regions ($\Delta \outlives \pi$). The result of a
\C{letregion} expression must have a type that is well-formed under a
context not containing the new region. This ensures that the value
obtained by evaluating a \C{letregion} expression contains no
references to the temporary objects inside the region.

The rule for \C{open} expression, unlike the rule for \C{letregion},
does not introduce any outlives relationship between the newly opened
region and any pre-existing region while checking the type of the
expression ($e$) under \C{open}. This prevents new objects allocated
inside the transferable region from storing references to those
outside.  Environment ($\env$) is extended with binding for the type
of root object while typechecking $e$.

The type rule for lambda expression typechecks the lambda-bound
expression ($e$) under an extended type environment containing
bindings for function's arguments, assuming that region parameters are
live, and that declared constraints over region parameters hold. The
constraints ($\phi$) are required to be well-formed under $\Delta$
extended with the function's region parameters ($\rho\rhobar$). Note
that the closure is always allocated in the current allocation context
($\rgn$). This prevents the closure from escaping the context in which
it is created, thus trivially ensuring the safety of any dereferences
inside the closure.

% The type rule for the \C{new} expression is a straightforward
% adaptation of corresponding FGJ rule to the the region type setting.

>>>>

The purpose of \name's region type system is to enforce the key
invariant required for memory safety, namely that an object $o_1$ in a
region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
$R_2$ is guaranteed to outlive $R_1$.  While the invariant is easily
stated, enforcing it in the presence of first-class dynamic regions,
parametric polymorphism (generics), and higher-order functions
requires new reasoning principles that we formally develop in this
section. We introduce \fbname (\FB), an explicitly typed core language
(with region types) that incorporates the features introduced in the
previous section. \fbname builds on the Featherweight Generic Java
(FGJ)~\cite{fgj} formalism, and reuses notations and various
definitions from~\cite{fgj}, such as the definition of type
well-formedness for the core (region-free) language.  (The language
used in Section~\ref{sec:overview} is essentially a version of FGJ
without region types and with some syntactic sugar.)

@@

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$. Class fields are not allowed to
be type polymorphic, in keeping with C\#, which constitutes the core
of \name.

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote objects contained in a single region, it
yeilds benefits that far outweigh the costs. In particular, it lets us
support region-polymorphic higher-order functions as class fields.
This allows, for example, a generic class, whose type parameters are
$a$ and $b$, to contain a region-polymorphic function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

Like classes, methods can also exhibit constrained region
polymorphism.  A method definition in \FB is necessarily polymorphic
over its allocation context (\S~\ref{sec:alloc-ctxt}), and optionally
polymorphic with respect to the regions containing its arguments
(i.e., a method has at least one region parameter). Region
parameters, like those on classes, are qualified with constraints
($\phi$).
% Region parameters on the methods, like those on classes, are
% accompanied by constraints ($\phi$) capturing the conditions that the
% parameters need to satisfy for the method to be considered well-formed
% (Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
% or $\rho^a_m$) is the first and inevitable region parameter of every
% method in \FB. 
If a method is not intended to be polymorphic with respect to its
allocation context (for example, if its allocation context needs to be
same as the allocation region of its \emph{this} argument), then the
required monomorphism can be captured as an equality constraint in
$\phi$.

% Like methods, functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
\FB extends FGJ's expression language with a lambda expression and an
application expression ($e\inang{\rbar}(\bar{e})$) to define
and apply functions (lambdas). Functions, like methods, exhibit
constrained region polymorphism, as evident in their arrow region type
($\inang{\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
\tau$).
% at the prenex position, similar to ML type schemes. However,
% unlike ML, we don't distinguish between types and type schemes; any of
% the $\tau$'s in the arrow type can themselves be region-parametric
% arrow types. Like with methods, region parameters on functions can be
% instantiated when they are applied.
% In this respect, our region type system is more like
% System F's type system, which admits higher-rank parametric
% polymorphism. Like System F, \FB provides a region instantiation
% expression ($e\inang{\ralloc\rbar}$), and a region generalization
% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
% instantiate and generalize region variables. 
A function, like a method, is necessarily polymorphic w.r.t its
allocation context.  Since a function closure can escape the context
in which it is created, it is important to keep track of the region in
which it is created in order to avoid unsafe dereferences. The $\rgn$
annotation above the arrow in the arrow type denotes the allocation
region of the corresponding closure. 
% Note that it is important to distinguish
% between the allocation context argument ($\rhoalloc$) of a function
% and the allocation region ($\rgn$) of its closure. In \name, the later
% corresponds to the region where a \C{Func} object is allocated, while
% the former corresponds to the region where it is applied
% For instance, in the following example:
% \begin{codejava}
% letregion $\rgn$ {
%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
%   in f
% }
% \end{codejava}
% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
% in $\rgn$, and (b). when executed under an allocation context
% $\rho^a$, the closure returns an object allocated in $\rho^a$.

@@

% Transferable regions are first-class objects in \name, hence in \FB.
% The type of a transferable region handler is of the form
% $\RgnZT{\rgn}$, where, like any other object, $\rgn$ denotes the
% region in which the handler object is allocated. However, region
% handlers are often required to escape the scope and address

@@

The type rules distinguish between the \C{new} expressions that create
objects of  the \RgnZ class, and \C{new} expressions that create
objects of other classes. The rule for the latter relies on an
auxiliary definition called $\fields$ (undefined for $\RgnZ$ class)
that returns the sequence of type bindings for fields (instance
variables) of a given class type.  Like in FGJ, the names and types of
a constructor's arguments in \FB are same as the names and types of
its class's fields.

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\pi \notin \Delta$). This condition is needed in order
to prevent the new region from incorrectly assuming existing outlives
relationships on an eponymous region. The expression ($e$) under
\C{letregion} is typechecked with the new region as its allocation
context, under the assumption that the region is live ($\Delta \cup
\{\pi\}$) and that it is outlived by all existing live regions
($\Delta \outlives \pi$). The result of a \C{letregion} expression
must have a type that is well-formed under a context not containing
the new region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region.

The rule for \C{open} expression, unlike the rule for \C{letregion},
does not introduce any outlives relationship between the newly opened
region and any pre-existing region while checking the type of the
expression ($e$) under \C{open}. This prevents new objects allocated
inside the transferable region from storing references to those
outside. The newly opened region becomes the allocation context for
$e$, which is type checked under an environment ($\env$) extended with
the type binding for the root object.

The type rule for lambda expression typechecks the lambda-bound
expression ($e$) under an extended type environment containing
bindings for function's arguments, assuming that region parameters are
live, and that declared constraints over region parameters hold. The
constraints ($\phi$) are required to be well-formed under $\Delta$
extended with the function's region parameters ($\rho\rhobar$).
Unlike a typical object, a function closure might capture references
that may become unsafe if the closure escapes its allocation context.
\FB prevents this scenario by requiring the function closure to always
be allocated in the current allocation context ($\rgn$).

The type rule for method application uses auxiliary definitions
$\mtype$ and $\bound$ which originate from the FGJ calculus - the
former gives the type of a class method and the latter defines the
bound of a type variable. Method application involves instantiation of
method's region parameters, and the type rule requires the first
region parameter to be instantiated with the current allocation
context ($r$). The type rule for function application does similarly.
This requirement ensures the safety of closures as demonstrated by the
following example.

Consider a method $m$ with two region parameters - $\rho_0$ and
$\rho_1$ (i.e., $m\inang{\rho_0,\rho_1}(\dots)$). Suppose the method
immediately returns a function closure that contains a reference to
(an object in) $\rho_1$. Since function closures are always allocated
in the current allocation context, the closure is allocated in
$\rho_0$, the allocation context for the method body\footnote{
Fig.~\ref{fig:fb-staticsem-1} of the Appendix contains the method
well-formedness rule, which is on the similar lines as the lambda type
rule shown in Fig.~\ref{fig:fb-staticsem}}. Now, consider the
following use of the method:
\begin{codejava}
letregion R0 in
  let f = letregion R1 in m<R0,R1>()
  in f()
\end{codejava}
Observe that the first region parameter of \C{m} is instantiated with
\C{R0}, while the allocation context for the method call is \C{R1}.
The function \C{f} returned by \C{m} stores a reference to \C{R1},
which is unsafe when \C{f} is finally called. The type system
fortunately disallows this scenario by enforcing certain restrictions
during method and function applications as described above.
% The type rule for the \C{new} expression is a straightforward
% adaptation of corresponding FGJ rule to the the region type setting.
